---
title: "Statistical Inference - Inferential data analysis"
author: "Alexandre Georges"
output: pdf_document
---

# Loading and basic analyses

## Dataset

*Basic information about the Tooth Growth dataset*

```{r, echo = TRUE}
head(ToothGrowth)
str(ToothGrowth)
summary(ToothGrowth)
var(ToothGrowth$len)
hist(ToothGrowth$len, col = 'red', xlab = 'Tooth length', main = 'Histogram of tooth length')
```

**The tooth length's variance is high and thus length is quite spreaded over a big interval: from 5 to 30.**

```{r, include = FALSE}
library(dplyr)
```

## Orange juice supplement

*Tooth length mean and variance for orange juice*

```{r, echo = TRUE}
OJToothGrowth <- ToothGrowth[ToothGrowth$supp == 'OJ', ]
summary(OJToothGrowth)
var(OJToothGrowth$len)
hist(OJToothGrowth$len, col = 'red', xlab = 'Tooth length', main = 'Histogram of tooth length for orange juice')
```

**There is a big drop of the tooth length for the interval 10 to 15.**

*Mean and variance grouped by dosage*

```{r, echo = TRUE}
summarize(group_by(OJToothGrowth, dose), mean = mean(len), variance = var(len))
```

**The variance decreases and the tooth length increases when the dosage increases.**

## Vitamin C supplement

*Tooth length mean and variance for vitamin C*

```{r, echo = TRUE}
VCToothGrowth <- ToothGrowth[ToothGrowth$supp == 'VC', ]
summary(VCToothGrowth)
var(VCToothGrowth$len)
hist(VCToothGrowth$len, col = 'red', xlab = 'Tooth length', main = 'Histogram of tooth length for vitamin C')
```

**The tooth length is more spreaded than with orange juice and there is no noticeable difference between the intervals.**

*Mean and variance grouped by dosage*

```{r, echo = TRUE}
summarize(group_by(VCToothGrowth, dose), mean = mean(len), variance = var(len))
```

**The variance is high with a dosage of 2 compared to the other dosages and the tooth length increases when the dosage increases.**

# Basic summary of the data

A good summary of the data was provided in the section above. To complete this basic analysis, the charts below show how spreaded the tooth length is.

## Comparison of orange juice and vitamin C

```{r, echo = TRUE}
library(ggplot2)
ggplot(ToothGrowth, aes(x = factor(supp), y = len, fill = supp)) +
  geom_boxplot() +
  labs(x = 'Supplement', y = 'length')
```

## Comparison of the supplements and dosage.

```{r, echo = TRUE}
ggplot(ToothGrowth, aes(x = factor(dose), y = len, fill = supp)) +
  geom_boxplot() +
  facet_grid(. ~ supp) +
  labs(x = 'Dose', y = 'length')
```

# Tests to compare tooth growth by supp and dose

## Test for orange juice and vitamin C

*T test between orange juice and vitamin C*

```{r, echo = TRUE}
t.test(OJToothGrowth$len, VCToothGrowth$len, var.equal = TRUE)
```

**The P value is small but slightly above 5%. So we can not tell for sure if there is a difference between the 2 groups. However the 95% confidence interval tends to show higher tooth growth for the orange juice than for the vitamin C.**

## Tests between dosages

*T tests between dosages for both supplements*

### Orange juice

```{r, echo = TRUE}
OJToothGrowth05 <- OJToothGrowth[OJToothGrowth$dose == 0.5, ]
OJToothGrowth1 <- OJToothGrowth[OJToothGrowth$dose == 1, ]
OJToothGrowth2 <- OJToothGrowth[OJToothGrowth$dose == 2, ]
```
```{r, echo = TRUE}
t.test(OJToothGrowth05$len, OJToothGrowth1$len, var.equal = TRUE)
```
```{r, echo = TRUE}
t.test(OJToothGrowth05$len, OJToothGrowth2$len, var.equal = TRUE)
```
```{r, echo = TRUE}
t.test(OJToothGrowth1$len, OJToothGrowth2$len, var.equal = TRUE)
```

**The P value is very small, the dosage has a big impact on the length (higher the dosage, higher the tooth length). The 95% confidence interval shows a big gap between dosages. However the difference is less pronounced for the dosages 1 and 2.**

### Vitamin C

```{r, echo = TRUE}
VCToothGrowth05 <- VCToothGrowth[VCToothGrowth$dose == 0.5, ]
VCToothGrowth1 <- VCToothGrowth[VCToothGrowth$dose == 1, ]
VCToothGrowth2 <- VCToothGrowth[VCToothGrowth$dose == 2, ]
```
```{r, echo = TRUE}
t.test(VCToothGrowth05$len, VCToothGrowth1$len, var.equal = TRUE)
```
```{r, echo = TRUE}
t.test(VCToothGrowth05$len, VCToothGrowth2$len, var.equal = TRUE)
```
```{r, echo = TRUE}
t.test(VCToothGrowth1$len, VCToothGrowth2$len, var.equal = TRUE)
```

**The P value is very small, the dosage has a big impact on the length (higher the dosage, higher the tooth length). The 95% confidence interval shows a big gap between dosages.**

# Conclusion

## Supplements

The comparison between the supplements does not show any obvious difference, the orange juice according to means is slightly more effective.

## Dosage

For both supplements, the p values are small and the 95% confidence intervals are big therefore the dosage has a big impact on the tooth length (higher the dosage, higher the tooth length). This result was already visible on the box plot and is now confirmed.
